\bf Title here

### 2020-08-04 paper/paper.tex
update paper.tex
minor corrections

\bigskip
\begin{abstract}
Friedman \& Tukey commented on their intial paper on projection pursuit
in 1974 that ``the technique use for maximising the projection index
strongly influences both the statistical and the computational aspects
of the procedure.'' While many projection pursuit indices have been
proposed in the literature, few concerns the optimisation procedure. In
this paper, we developed a system of diagnostics aiming to visually
learn how the optimisation procedures find its way towards the optimum.
This diagnostic system can be applied to more general to help
practitioner to unveil the black-box in randomised iteartive
(optimisation) algorithms. An R package, ferrn, has been created to
implement this diagnostic system.
\end{abstract}

@@

In an optimization problem the goal is to find the best solution within
the space of all feasible solutions which typically is represented by a
set of constraints. The problem consists on optimizing an objective
function \(f: S \rightarrow \Re\) with \(S \in \Re^n\) in a reduced
spaced given by the problem constraints to either minimize or maximize a
function\ldots{}. Gradient based optimization has been typically used to
solve such problems. However, there are many situations where
derivatives of an objective function are unavailable or unreliable and
therefore traditional methods based on derivatives are not the best
option to solve an optimization problem.

Derivative free methods provides another option to optimise the
objective function without evaluating any gradient or Hessian
information and a particular class of methods: direct search, has gained
its popularity through its conceptual simplicity. However, the whole
searching process in the algorithm remains a black-box. Plots are
usually used to evaluate and compare the performance of different
algorithms but it can easily become tedious because the code will have
to be modified significantly when comparing different parameters in the
algorithms. For example, a categorical variable with 5 levels can be
easily mapped onto colour while mapping another categorical variable
with 30 levels will not make the plot informative. Thus the plot needs
to be re-designed to better suits the characteristics of the parameter
(whether the parameter is a scalar or a matrix? whether the parameter is
quantitative or categorical? If categorical, how many levels does the
parameter have?). This motivates the design of a visual diagnostic
framework for optimisation algorithms based on the idea of a
\emph{global object}.

The remainder of the paper is organised as follows. Section \ref{DFO}
gives a general overview of optimisation methods, specifically
derivative free optimisation. Section \ref{tour} introduces the
optimisation problem in a specific problem: projection pursuit guided
tour and existing derivative free algorithms for solving it. Section
\ref{vis-diag} and \ref{vis-diag-plots} together present the new
construction of a visual diagnostic framework that unveil the black-box
of optimisation. Section \ref{vis-diag} focuses on motivational ideas
and data structure and Section \ref{vis-diag-plots} provides definitions
of different diagnostic plots using grammar of graphic and examples.
Finally, \ref{implementation} mentions the R package: ferrn that
implements the visual diagnostics described above.

@@

Given an objective function \(f\), one way of optimising it is to equate
its gradient to zero. In modern optimisation problems, gradient
information can be hard to evaluate or sometimes even impossible and
Derivative-Free Optimisation (DFO) methods can be useful to approach
these problems. One common class of methods in DFO is
\emph{Direct-search methods}. Coined by \citet{hooke1961direct}, direct
search methods don't require any gradient or Hessian information and has
gained its popularity through its simplicity in use and reliability in
solving complicated practical problems. Depends on whether a random
sample is used in the search, this class of methods can be further
classified as \emph{stochastic} or \emph{deterministic}. The stochastic
version of the direct-search methods will be the main optimisation
procedure analysed in this paper.

@@

{[}Are we using projection pursuit/guided tour to better understand the
convergence of optimization algorithms visually in combination with the
algorithms discussed below? Or we are focusing on the optimisation
problem only within the project pursuit context? Some of the problems
listed below are also applicable to optimization problem in general too.
ppp{]}

Below listed several issues in projection pursuit optimisation. Some are
general optimisation problems, while others are more specific for PP
optimisation.

\begin{itemize}
\item
  \emph{Finding global maximum}: Although finding local maximum is
  relatively easy with developed algorithms, it is generally hard to
  guarantee global maximum in a problem where the objective function is
  complex or the number of decision variables is large. Also, there are
  discussions on how to avoid getting trapped in a local optimal in the
  literature.
\item
  \emph{optimising non-smooth function}: When the objective function is
  non-differentiable, derivative information can not be obtained, which
  means traditional gradient- or Hessian- based methods are not
  feasible. Stochastic optimisation method could be an alternative to
  solve these problems.
\item
  \emph{computation speed}: The optimisation procedure needs to be fast
  to compute since tours produces real-time animation of the projected
  data.
\item
  \emph{consistency result in stochastic optimisation}: In stochastic
  algorithm, researchers usually set a seed to ensure the algorithm
  producing the same result for every run. This practice supports
  reproducibility, while less efforts has been made to guarantee
  different seeds will provide the same result.
\item
  \emph{high-dimensional decision variable}: In projection pursuit, the
  decision variable includes all the entries in the projection matrix,
  which is high-dimensional. Researcher would be better off if they can
  understand the relative position of different projection matrix in the
  high-dimensional space.
\item
  \emph{role of interpolation in PP optimisation}: An optimisation
  procedure usually involves iteratively finding projection bases that
  maximises the index function, while tour requires geodesic
  interpolation between these bases to produce a continuous view for the
  users. It would be interesting to see if the interpolated bases could,
  in reverse, help the optimisation reach faster convergence.
\end{itemize}

@@

Coined by \citet{friedman1974projection}, projection pursuit detects
interesting structures (i.e.~clustering, outliers and skewness) in
multivariate data via projecting onto lower dimensions. Let
\(\mathbf{X}_{n \times p}\) be the data matrix, an n-d projection can be
seen as a linear transformation \(T: \mathbb{R}^p \mapsto \mathbb{R}^d\)
defined by \(\mathbf{P} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{P}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis. Define
\(f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}\) be an index function
that maps the projection basis \(\mathbf{A}\) onto an index value \(I\),
this function is commonly known as the projection pursuit index (PPI)
function, or the index function and is used to measure the
``interestingness'' of a projection. A number of index functions have
been proposed in the literature to detect different data structures,
including Legendre index \citep{friedman1974projection}, Hermite index
\citep{hall1989polynomial}, natural Hermite index
\citep{cook1993projection}, chi-square index
\citep{posse1995projection}, LDA index \citep{lee2005projection} and PDA
index \citep{lee2010projection}.

In their initial paper, \citet{friedman1974projection} noted that
``\ldots{}, the technique use for maximising the projection index
strongly influences both the statistical and the computational aspects
of the procedure.'' Hence, effective optimisation algorithms are
necessary for projection pursuit to find the bases that give interesting
projections. While we leave the formal construction of the optimisation
problem and existing algorithms in section \ref{tour-optim}, we outlines
the general idea here. Given a random starting (current) basis,
projection pursuit repetitively searches for candidate bases nearby
until it finds one with higher index value than the current basis. In
the second round, that basis becomes the current basis and the
repetitive sampling continues. The process ends until no better basis
can be found or one of the termination criteria is reached.

@@

\citet{posse1995projection} proposed a random search algorithm that
samples a candidate basis \(\mathbf{A}_{l}\) in the neighbourhood of the
current basis \(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}\),
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{rand}\) is a randomly generated matrix with the same
dimension as \(\mathbf{A}_{\text{cur}}\). The optimiser keeps sampling
bases near the current basis until it finds one with higher index value
than the current basis and then outputs it for guided tour to construct
the interpolation path. A new round of search continues to find a better
basis after the interpolation finishes. The halfing parameter \(c\) with
default value of 30 is designed to adjust the searching neighbourhood
\(\alpha\). When the search needs to sample more than \(c\) number of
basis to find an outputted basis, the neighbourhood parameter \(\alpha\)
will be reduced by half in the next iteration. If the optimiser can't
find a better basis within the maximum number of tries \(l_{\max}\), the
algorithm stops. The algorithm is summarised in Algorithm
\ref{random-search} for one iteration.

@@

Simulated annealing
\citep[\citet{kirkpatrick1983optimization}]{bertsimas1993simulated} is a
non-derivative procedure based on a non-increasing cooling scheme
\(T(l)\). Given an initial \(T_0\), the temperature at iteration \(l\)
is defined as \(T(l) = \frac{T_0}{\log(l + 1)}\). The simulated
annealing algorithm works as follows. Given a neighbourhood parameter
\(\alpha\) and a randomly generated orthonormal basis \(A_{rand}\), a
candidate basis is sampled by
\(\mathbf{A}_{t+1} = (1 - \alpha)\mathbf{A}_{t} + \alpha \mathbf{A}_{rand}\).
If the index value of the candidate basis is larger than the one of the
current basis, the candidate basis is outputted. While if it is smaller,
the candidate has another chance of being accepted with probability
\(P= \min\left\{\exp\left[\frac{I_{cur} - I_{l}}{T(l)}\right],1\right\}\)
where \(I_{\cdot}\) denotes the index value of a given basis. The
algorithm can be summarised as in Algorithm \ref{simulated_annealing}.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$; Initial temperature: $T_0$}
    \Output{The global object; The target basis: $\mathbf{A}_{l}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = (1 - \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}$ ensuring $\mathbf{A}_{l}$ is orthonormal \;
    Compute $I_{l} = f(\mathbf{A}_{l})$ and $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I_{l} > I_{\text{cur}}$}{
        \KwRet{$\mathbf{A}_{l}$} \;
      }{
        Compute $P= \min\left\{\exp\left[\frac{I_{\text{cur}} -I_{l}}{T(l)}\right],1\right\}$\;
        Draw $D$ from a uniform distribution: $D \sim \text{Unif(0, 1)}$\;
        \If{$P > D$}{
           \KwRet{$\mathbf{A}_{l}$} \;
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

@@

Visualisation has been widely used in exploratory data analysis.
Presenting information in a graphical format often allows people to see
information they would otherwise not seen. This motivates us \textbf{to
develop a visual framework to diagnose optimisation algorithms in
projection pursuit with a aim to understand and compare different
existing algorithms.} This study is meaningful because \textbf{as more
complex index functions are proposed based on statistical theory, only
computationally effective optimisation algorithms allow us to find the
projection basis that presents interesting structure in the data}.

The idea of generalised framework for diagnostic plots is inspired by
the concept of grammar of graphic \citep{wickham2010layered}, which
powers the primary graphical system in R, ggplot2 \citep{ggplot2}. In
grammar of graphic, plots are not defined by its appearance
(i.e.~boxplot, histogram, scatter plot etc) but by ``stacked layers''.
Using this design, ggplot doesn't have to develop a gazillion of
functions that each produces a different type of plot from a different
data structure. Instead, it aesthetically maps variables (and its
statistical transformation) in a dataset to different geometric objects
(points, lines, box-and-whisker etc) and builds the plot through
overlaying different layers.

@@

We simulate some random variables of size 1000 with different
structures. \texttt{x1}, \texttt{x8}, \texttt{x9} and \texttt{x10} are
simulated from normal distribution with zero mean and variance of one as
in equation \ref{eq:sim-norm}. When using projection pursuit to explore
the data structure based on its departure from normality, the entry in
the projection basis for these variables should be close to zero in
theory. \texttt{x2} to \texttt{x7} are mixture of normal distributions
with different weights and locations. Equation \ref{eq:sim-x2} to
\ref{eq:sim-x7} outlines the distribution where each variable is
simulated from and Figure \ref{origin-data} shows the histogram of each
variable except \texttt{x3}. All the variables are then scaled to have
variance of one before running the projection pursuit.

@@

We form our first dataset using variable \texttt{x1}, \texttt{x2},
\texttt{x8}, \texttt{x9} and \texttt{x10} and run the guided tour with
optimiser \texttt{search\_better}. When the optimisation ends, the
global object will be stored and printed (it can be turned off by
supplying argument \texttt{print\ =\ FALSE}). Additional messages during
the optimisation can be displayed by argument \texttt{verbose\ =\ TRUE}.
Below shows the first ten rows of the global object. Notice that the
tibble object allows the list-column \texttt{basis} to be printed out
nicely with the dimension of the projection basis readily available.

@@

A primary interest of diagnosing an optimisation algorithm is to study
how it finds its optimal progressively. We could plot the index value
across its natural ordering, however, different iterations may have
different number of points and, towards the end of the search there
could easily be hundreds of bases being tested before the target basis
is found. In the plot, points from those iterations towards the end will
occupy the vast majority of the plot space. This motivates to use
summarisation. Rather than knowing the index value of \emph{every}
basis, we are more interested to have a general summary of all the index
value in that iteration and more importantly, the basis with the largest
index value (since it prescribes the next geodesic interpolation and
future searches).

Boxplot is a suitable candidate that provides five points summary of the
data, however it has one drawback: it doesn't report the number of point
in each box. We may risk losing information on how many points it takes
to find the target basis by displaying the boxplot alone for all
\texttt{tries}. Thus, the number of point in each iteration is displayed
at the bottom of each box and we provide options to switch iteration
with small number of points to a point geometry, which is achieved via
an \texttt{cutoff} argument. A line geometry is also added to link the
points with the largest index value in each iteration. This helps to
visualise the improvement made in each iteration. Using the concept of
\emph{gramma of graphics} \citep{wickham2010layered}, the plot for
exploring index value can be defined in three layers as:

@@

We choose variable \texttt{x1}, \texttt{x2}, \texttt{x3}, \texttt{x8},
\texttt{x9} and \texttt{x10} to perform a 2D projection with tour.
Parameter \texttt{search\_f\ =\ tour::search\_better} and
\texttt{max.tries\ =\ 500} is used. The index value of the searching
points are shown in Figure \ref{points}. Label at the bottom indicates
the number of observations in each iteration and facilitates the choice
of \texttt{cutoff} argument (by default \texttt{cutoff\ =\ 15}). We
learn that the \texttt{search\_better} quickly finds better projection
basis with higher index value at first and the takes longer to find a
better one later.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/points-tries-1.pdf}
\caption{\label{points}A comparison of plotting the same search points
with different plot designs. The left plot doesn't efficiently use the
plot space to convey information from the plot while the right plot
provides good summarisation of data and number of points in each tries.}
\end{figure}

@@

Sometimes, rather than explore the searching points, we may be
interested in explore the points on the interpolation path (target and
interpolating points) since these points will be played by the tour
animation. Since interpolation paths are geodesically the shortest, a
summarisation using boxplot geometry is no longer needed. The slightly
modified plot definition is shown below:

@@

We use the same dataset as the toy example above to explore the search
function \texttt{search\_better} and we want to learn how the index
value changes on the interpolation path for \texttt{holes} index. From
the left panel of Figure \ref{interruption}, we observe that when
interpolating from the current basis to the target basis, the index
value may not be monotone: we could reach a basis with higher index
value than the target basis on the interpolation path. In this sense, we
would be better off using the basis with the highest index value on the
interpolation path as the current basis for the next iteration (rather
than using the target basis). Hence, an interruption is constructed to
accept the interpolating bases only up to the one with the largest index
value. After implementing this interruption, the search finds higher
final index value with fewer steps as shown in the right panel of Figure
\ref{interruption}.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/interruption-1.pdf}
\caption{\label{interruption}Trace plots of the interpolated basis with
and without the interruption. The interruption stops the interpolation
when the index value starts to decrease at id = 60. The implementation
of the interuption finds an ending basis with higher index value using
fewer steps.}
\end{figure}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  the chosen basis can't be too close to the current basis
\item
  the percentage improvement of the index value can't be too small
\item
  the searching neighbourhood can't be too small
\item
  the number of iteration can't exceed the \texttt{max.tries}
\end{enumerate}

@@

Slight variation of plot definition due to the addition of polishing
points is as follows:

@@

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/polish-1.pdf}
\caption{\label{trace-compare}Breakdown of index value when using
different max.tries in search better in conjunction with search polish.
Both attain the same final index value after the polishing while using a
max.tries 25 is not sufficient to find the ture maximum.}
\end{figure}

@@

\emph{Example: understand search\_geodesic} {[}feel like this example is
merely explaining search geodesic algorithm, so maybe introduce the
animated plot here? xxx{]} \texttt{search\_geodesic} is a two-stage
ascending algorithm with four different stages in the search and a PCA
plot useful to understand how the algorithm progresses and the relative
position of each basis in the PCA projected 2D space. Starting from the
start basis, a directional search is conducted in a narrow neighbourhood
on five random directions. The best one is picked and a line search is
then run on the geodesic direction to find the target basis. The
starting and target bases are then interpolated. In the next iteration,
the target basis becomes the current basis and then procedures
continues.

@@

\[\max \left[F_{\mathbf{P}}(u) - F_{y}(u)\right]\]

can be seen as a function of the projection matrix
\(\mathbf{A}_{p \times 1}\) and hence a valid index function.

@@

Now we enlarge the dataset to include two informative variables:
\texttt{x2} and \texttt{x3} and remain 1D projection. In this case, two
local maximum appear when projection matrix being \([0, 1, 0, 0, 0, 0]\)
and \([0, 0, 1 ,0, 0, 0]\).

>>>>

\bigskip
\begin{abstract}
Friedman \& Tukey commented on their initial paper on projection pursuit
in 1974 that ``the technique used for maximising the projection index
strongly influences both the statistical and the computational aspects
of the procedure.'' While many projection pursuit indices have been
proposed in the literature, few concerns the optimisation procedure. In
this paper, we developed a system of diagnostics aiming to visually
learn how the optimisation procedures find its way towards the optimum.
This diagnostic system can be applied more generally to help
practitioner to unveil the black-box in randomised iterative
(optimisation) algorithms. An R package, ferrn, has been created to
implement this diagnostic system.
\end{abstract}

@@

In an optimization problem the goal is to find the best solution within
the space of all feasible solutions which typically is represented by a
set of constraints. The problem consists in optimizing an objective
function \(f: S \rightarrow \Re\) with \(S \in \Re^n\) in a reduced
spaced given by the problem constraints to either minimize or maximize a
function\ldots{}. Gradient based optimization has been typically used to
solve such problems. However, there are many situations where
derivatives of an objective function are unavailable or unreliable and
therefore traditional methods based on derivatives are not the best
option to solve an optimization problem.

Derivative free methods provide another option to optimise the
objective function without evaluating any gradient or Hessian
information and a particular class of methods, direct search, has gained
popularity through its conceptual simplicity. However, the whole
searching process in the algorithm remains a black-box. Plots are
usually used to evaluate and compare the performance of different
algorithms but it can easily become tedious because the code will have
to be modified significantly when comparing different parameters in the
algorithms. For example, a categorical variable with 5 levels can be
easily mapped onto colour while mapping another categorical variable
with 30 levels will not make the plot informative. Thus the plot needs
to be re-designed to better suit the characteristics of the parameter
(whether the parameter is a scalar or a matrix? whether the parameter is
quantitative or categorical? If categorical, how many levels does the
parameter have?). This motivates the design of a visual diagnostic
framework for optimisation algorithms based on the idea of a
\emph{global object}.

The remainder of the paper is organised as follows. Section \ref{DFO}
gives a general overview of optimisation methods, specifically
derivative free optimisation. Section \ref{tour} introduces the
optimisation problem in a specific problem: projection pursuit guided
tour and existing derivative free algorithms for solving it. Section
\ref{vis-diag} and \ref{vis-diag-plots} together present the new
construction of a visual diagnostic framework that unveil the black-box
of optimisation. Section \ref{vis-diag} focuses on motivational ideas
and data structure and Section \ref{vis-diag-plots} provides definitions
of different diagnostic plots using grammar of graphic and examples.
Finally, Section \ref{implementation} mentions the R package: ferrn that
implements the visual diagnostics described above.

@@

Given an objective function \(f\), one way of optimising it is to equate
its gradient to zero. In modern optimisation problems, gradient
information can be hard to evaluate or sometimes even impossible and
Derivative-Free Optimisation (DFO) methods can be useful to approach
these problems. One common class of methods in DFO is
\emph{Direct-search methods}. Coined by \citet{hooke1961direct}, direct
search methods don't require any gradient or Hessian information and has
gained its popularity through its simplicity in use and reliability in
solving complicated practical problems. Depending on whether a random
sample is used in the search, this class of methods can be further
classified as \emph{stochastic} or \emph{deterministic}. The stochastic
version of the direct-search methods will be the main optimisation
procedure analysed in this paper.

@@

{[}Are we using projection pursuit/guided tour to better understand the
convergence of optimization algorithms visually in combination with the
algorithms discussed below? Or we are focusing on the optimisation
problem only within the projection pursuit context? Some of the problems
listed below are also applicable to optimization problem in general too.
ppp{]}

Below are listed several issues in projection pursuit optimisation. Some are
general optimisation problems, while others are more specific to PP
optimisation.

\begin{itemize}
\item
  \emph{Finding global maximum}: Although finding local maximum is
  relatively easy with developed algorithms, it is generally hard to
  guarantee global maximum in a problem where the objective function is
  complex or the number of decision variables is large. Also, there are
  discussions on how to avoid getting trapped in a local optimum in the
  literature.
\item
  \emph{optimising non-smooth function}: When the objective function is
  non-differentiable, derivative information can not be obtained, which
  means traditional gradient- or Hessian- based methods are not
  feasible. Stochastic optimisation method could be an alternative to
  solve these problems.
\item
  \emph{computation speed}: The optimisation procedure needs to be fast
  to compute since tours produce real-time animation of the projected
  data.
\item
  \emph{consistency result in stochastic optimisation}: In stochastic
  algorithm, researchers usually set a seed to ensure the algorithm
  produces the same result for every run. This practice supports
  reproducibility, while less efforts has been made to guarantee
  different seeds will provide the same result.
\item
  \emph{high-dimensional decision variable}: In projection pursuit, the
  decision variable includes all the entries in the projection matrix,
  which is high-dimensional. Researchers would be better off if they could
  understand the relative position of different projection matrix in the
  high-dimensional space.
\item
  \emph{role of interpolation in PP optimisation}: An optimisation
  procedure usually involves iteratively finding projection bases that
  maximise the index function, while tour requires geodesic
  interpolation between these bases to produce a continuous view for the
  users. It would be interesting to see if the interpolated bases could,
  in reverse, help the optimisation reach faster convergence.
\end{itemize}

@@

Coined by \citet{friedman1974projection}, projection pursuit detects
interesting structures (i.e.~clustering, outliers and skewness) in
multivariate data via projecting onto lower dimensions. Let
\(\mathbf{X}_{n \times p}\) be the data matrix, an n-d projection can be
seen as a linear transformation \(T: \mathbb{R}^p \mapsto \mathbb{R}^d\)
defined by \(\mathbf{P} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{P}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis. Define
\(f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}\) to be an index function
that maps the projection basis \(\mathbf{A}\) onto an index value \(I\),
this function is commonly known as the projection pursuit index (PPI)
function, or the index function and is used to measure the
``interestingness'' of a projection. A number of index functions have
been proposed in the literature to detect different data structures,
including Legendre index \citep{friedman1974projection}, Hermite index
\citep{hall1989polynomial}, natural Hermite index
\citep{cook1993projection}, chi-square index
\citep{posse1995projection}, LDA index \citep{lee2005projection} and PDA
index \citep{lee2010projection}.

In their initial paper, \citet{friedman1974projection} noted that
``\ldots{}, the technique used for maximising the projection index
strongly influences both the statistical and the computational aspects
of the procedure.'' Hence, effective optimisation algorithms are
necessary for projection pursuit to find the bases that give interesting
projections. While we leave the formal construction of the optimisation
problem and existing algorithms to section \ref{tour-optim}, we outline
the general idea here. Given a random starting (current) basis,
projection pursuit repeatedly searches for candidate bases nearby
until it finds one with higher index value than the current basis. In
the second round, that basis becomes the current basis and the
repetitive sampling continues. The process ends until no better basis
can be found or one of the termination criteria is reached.

@@

\citet{posse1995projection} proposed a random search algorithm that
samples a candidate basis \(\mathbf{A}_{l}\) in the neighbourhood of the
current basis \(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}\),
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{rand}\) is a randomly generated matrix with the same
dimension as \(\mathbf{A}_{\text{cur}}\). The optimiser keeps sampling
bases near the current basis until it finds one with higher index value
than the current basis and then outputs it for guided tour to construct
the interpolation path. A new round of search continues to find a better
basis after the interpolation finishes. The halving parameter \(c\) with
default value of 30 is designed to adjust the searching neighbourhood
\(\alpha\). When the search needs to sample more than \(c\) number of
basis to find an outputted basis, the neighbourhood parameter \(\alpha\)
will be reduced by half in the next iteration. If the optimiser can't
find a better basis within the maximum number of tries \(l_{\max}\), the
algorithm stops. The algorithm is summarised in Algorithm
\ref{random-search} for one iteration.

@@

Simulated annealing
\citep[\citet{kirkpatrick1983optimization}]{bertsimas1993simulated} is a
non-derivative procedure based on a non-increasing cooling scheme
\(T(l)\). Given an initial \(T_0\), the temperature at iteration \(l\)
is defined as \(T(l) = \frac{T_0}{\log(l + 1)}\). The simulated
annealing algorithm works as follows. Given a neighbourhood parameter
\(\alpha\) and a randomly generated orthonormal basis \(\mathbf{A}_{\text{rand}}\), a
candidate basis is sampled by
\(\mathbf{A}_{t+1} = (1 - \alpha)\mathbf{A}_{t} + \alpha \mathbf{A}_{\text{rand}}\).
If the index value of the candidate basis is larger than the one of the
current basis, the candidate basis is outputted. While if it is smaller,
the candidate has another chance of being accepted with probability
\(P= \min\left\{\exp\left[\frac{I_{cur} - I_{l}}{T(l)}\right],1\right\}\)
where \(I_{\cdot}\) denotes the index value of a given basis. The
algorithm can be summarised as in Algorithm \ref{simulated_annealing}.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$; Initial temperature: $T_0$}
    \Output{The global object; The target basis: $\mathbf{A}_{l}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = (1 - \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ ensuring $\mathbf{A}_{l}$ is orthonormal \;
    Compute $I_{l} = f(\mathbf{A}_{l})$ and $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I_{l} > I_{\text{cur}}$}{
        \KwRet{$\mathbf{A}_{l}$} \;
      }{
        Compute $P= \min\left\{\exp\left[\frac{I_{\text{cur}} -I_{l}}{T(l)}\right],1\right\}$\;
        Draw $U$ from a uniform distribution: $U \sim \text{Unif(0, 1)}$\;
        \If{$P > U$}{
           \KwRet{$\mathbf{A}_{l}$} \;
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

@@

Visualisation has been widely used in exploratory data analysis.
Presenting information in a graphical format often allows people to see
information they would otherwise not see. This motivates us \textbf{to
develop a visual framework to diagnose optimisation algorithms in
projection pursuit with the aim to understand and compare different
existing algorithms.} This study is meaningful because \textbf{as more
complex index functions are proposed based on statistical theory, only
computationally effective optimisation algorithms allow us to find the
projection basis that presents interesting structure in the data}.

The idea of generalised framework for diagnostic plots is inspired by
the concept of grammar of graphic \citep{wickham2010layered}, which
powers the primary graphical system in R, ggplot2 \citep{ggplot2}. In
grammar of graphic, plots are not defined by its appearance
(i.e.~boxplot, histogram, scatter plot etc) but by ``stacked layers''.
Using this design, ggplot does not have to develop a gazillion of
functions that each produces a different type of plot from a different
data structure. Instead, it aesthetically maps variables (and its
statistical transformation) in a dataset to different geometric objects
(points, lines, box-and-whisker etc) and builds the plot through
overlaying different layers.

@@

We simulate some random variables of size 1000 with different
structures. \texttt{x1}, \texttt{x8}, \texttt{x9} and \texttt{x10} are
simulated from normal distribution with zero mean and variance of one as
in equation \eqref{eq:sim-norm}. When using projection pursuit to explore
the data structure based on its departure from normality, the entry in
the projection basis for these variables should be close to zero in
theory. \texttt{x2} to \texttt{x7} are mixture of normal distributions
with different weights and locations. Equation \eqref{eq:sim-x2} to
\eqref{eq:sim-x7} outlines the distribution where each variable is
simulated from and Figure \ref{origin-data} shows the histogram of each
variable except \texttt{x3}. All the variables are then scaled to have
unit variance before running the projection pursuit.

@@

We form our first dataset using variables \texttt{x1}, \texttt{x2},
\texttt{x8}, \texttt{x9} and \texttt{x10} and run the guided tour with
optimiser \texttt{search\_better}. When the optimisation ends, the
global object will be stored and printed (it can be turned off by
supplying argument \texttt{print\ =\ FALSE}). Additional messages during
the optimisation can be displayed by argument \texttt{verbose\ =\ TRUE}.
Below shows the first ten rows of the global object. Notice that the
tibble object allows the list-column \texttt{basis} to be printed out
nicely with the dimension of the projection basis readily available.

@@

A primary interest of diagnosing an optimisation algorithm is to study
how it finds its optimum progressively. We could plot the index value
across its natural ordering, however, different iterations may have
different number of points and, towards the end of the search there
could easily be hundreds of bases being tested before the target basis
is found. In the plot, points from those iterations towards the end will
occupy the vast majority of the plot space. This motivates to use
summarisation. Rather than knowing the index value of \emph{every}
basis, we are more interested to have a general summary of all the index
value in that iteration and more importantly, the basis with the largest
index value (since it prescribes the next geodesic interpolation and
future searches).

Boxplot is a suitable candidate that provides five points summary of the
data, however it has one drawback: it does not report the number of points
in each box. We may risk losing information on how many points it takes
to find the target basis by displaying the boxplot alone for all
\texttt{tries}. Thus, the number of point in each iteration is displayed
at the bottom of each box and we provide options to switch iteration
with small number of points to a point geometry, which is achieved via
a \texttt{cutoff} argument. A line geometry is also added to link the
points with the largest index value in each iteration. This helps to
visualise the improvement made in each iteration. Using the concept of
\emph{gramma of graphics} \citep{wickham2010layered}, the plot for
exploring index value can be defined in three layers as:

@@

We choose variables \texttt{x1}, \texttt{x2}, \texttt{x3}, \texttt{x8},
\texttt{x9} and \texttt{x10} to perform a 2D projection with tour.
Parameter \texttt{search\_f\ =\ tour::search\_better} and
\texttt{max.tries\ =\ 500} is used. The index value of the searching
points are shown in Figure \ref{points}. The label at the bottom indicates
the number of observations in each iteration and facilitates the choice
of \texttt{cutoff} argument (by default \texttt{cutoff\ =\ 15}). We
learn that the \texttt{search\_better} quickly finds better projection
basis with higher index value at first and then takes longer to find a
better one later.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/points-tries-1.pdf}
\caption{\label{points}A comparison of plotting the same search points
with different plot designs. The left plot does not use the
plot space efficiently to convey information from the plot while the right plot
provides good summarisation of data and number of points in each tries.}
\end{figure}

@@

Sometimes, rather than exploring the searching points, we may be
interested in exploring the points on the interpolation path (target and
interpolating points) since these points will be played by the tour
animation. Since interpolation paths are geodesically the shortest, a
summarisation using boxplot geometry is no longer needed. The slightly
modified plot definition is shown below:

@@

We use the same dataset as the toy example above to explore the search
function \texttt{search\_better} and we want to learn how the index
value changes on the interpolation path for the \texttt{holes} index. From
the left panel of Figure \ref{interruption}, we observe that when
interpolating from the current basis to the target basis, the index
value may not be monotone: we could reach a basis with higher index
value than the target basis on the interpolation path. In this sense, we
would be better off using the basis with the highest index value on the
interpolation path as the current basis for the next iteration (rather
than using the target basis). Hence, an interruption is constructed to
accept the interpolating bases only up to the one with the largest index
value. After implementing this interruption, the search finds a higher
final index value with fewer steps as shown in the right panel of Figure
\ref{interruption}.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/interruption-1.pdf}
\caption{\label{interruption}Trace plots of the interpolated basis with
and without interruption. The interruption stops the interpolation
when the index value starts to decrease at id = 60. The implementation
of the interuption finds an ending basis with higher index value using
fewer steps.}
\end{figure}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  the chosen basis can't be too close to the current basis
\item
  the percentage improvement of the index value can't be too small
\item
  the searching neighbourhood can't be too small
\item
  the number of iteration can't exceed \texttt{max.tries}
\end{enumerate}

@@

A slight variation of the plot definition due to the addition of polishing
points is as follows:

@@

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/polish-1.pdf}
\caption{\label{trace-compare}Breakdown of index value when using
different max.tries in search better in conjunction with search polish.
Both attain the same final index value after the polishing while using a
\texttt{max.tries} of 25 is not sufficient to find the true maximum.}
\end{figure}

@@

\emph{Example: understand search\_geodesic} {[}feel like this example is
merely explaining search geodesic algorithm, so maybe introduce the
animated plot here? xxx{]} \texttt{search\_geodesic} is a two-stage
ascending algorithm with four different stages in the search and a PCA
plot useful to understand how the algorithm progresses and the relative
position of each basis in the PCA projected 2D space. Starting from the
start basis, a directional search is conducted in a narrow neighbourhood
on five random directions. The best one is picked and a line search is
then run on the geodesic direction to find the target basis. The
starting and target bases are then interpolated. In the next iteration,
the target basis becomes the current basis and then the procedures
continues.

@@

\[\max \left[F_{\mathbf{P}}(u) - F_{y}(u)\right]\]
can be seen as a function of the projection matrix
\(\mathbf{A}_{p \times 1}\) and hence a valid index function.

@@

Now we enlarge the dataset to include two informative variables:
\texttt{x2} and \texttt{x3} and remain 1D projection. In this case, two
local maxima appear with projection matrix being \([0, 1, 0, 0, 0, 0]\)
and \([0, 0, 1 ,0, 0, 0]\).
